---
title: "Clustering analysis"
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(ggbiplot)
library(cluster)
library(factoextra)
```

## k-means clustering

### k-means clustering, fixed k

Loading the banknotes dataset

```{r}
banknotes <- read_csv("data/banknotes.csv")
```

Preprocess the data by performing a PCA, and truncate the data to the first 3 PCs:

```{r}
bnotes_pca <- prcomp(select(banknotes, Length:Top), scale.=TRUE)
```

```{r}
bnotes_coords <- as.data.frame(bnotes_pca$x) |> select(PC1:PC3)
```

We know that the banknotes are either genuine or counterfeit, hence we assume there are 2 clusters

```{r}
bnotes_km <- kmeans(bnotes_coords, centers=2)
```

Plot the result of clustering on the first two PCA coordinates

```{r}
ggbiplot(bnotes_pca, scale=0, groups=as.factor(bnotes_km$cluster))
```

Compare with ground truth

```{r}
ggbiplot(bnotes_pca, scale=0, groups=as.factor(banknotes$Currency))
```

### k-means clustering, determining k

Try to determine the "natural" number of cluster by looking for an elbow in within-cluster-sum-of-squares

```{r}
fviz_nbclust(bnotes_coords, kmeans, method = "wss")
```

Now try to determine the number of cluster using Silhouette score instead:

```{r}
fviz_nbclust(bnotes_coords, kmeans, method = "silhouette")
```

Inspect the Silhouette plot when k = 2

```{r}
dists <- dist(bnotes_coords)
sil <- silhouette(bnotes_km$cluster, dists)

p <- fviz_silhouette(sil)
p$layers[[2]]$aes_params$colour <- "black"

print(p)
```

Compare with the case of k = 5

```{r}
n_clust <- 5

dists <- dist(select(banknotes, Length:Top))
clust <- kmeans(select(banknotes, Length:Top), n_clust)
sil <- silhouette(clust$cluster, dists)

p <- fviz_silhouette(sil)
p$layers[[2]]$aes_params$colour <- "black"

print(p)
```

