---
title: "Week 10 exercises"
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(broom)
library(mgcv)
library(gratia)
library(janitor)
```

## Exercise 1: Pelagic bioluminescence, take 1

In Gillibrand et al. (2007), pelagic bioluminescence along a depth gradient in the northeast Atlantic Ocean is analysed. The data is found in the file pelagic_bioluminescence.csv

```{r}
lumin <- read_csv("data/pelagic_bioluminescence.csv") |> clean_names()
```

```{r}
glimpse(lumin)
```

```{r}
ggplot(lumin, aes(x=sources, y=sample_depth)) + geom_point() + 
  scale_y_continuous(transform="reverse") + facet_wrap(~station)
```

1. Subset the dataset to include only Station 13, and fit an additive model for which the amount of bioluminescence (`Sources`) depends semi-parametrically on sample depth (`SampleDepth`). Display the resulting summary.

2. Visualize the dataset and the fit (including its uncertainty). Use the usual oceanographic convention that depth is on the vertical axis and increases downwards.


## Exercise 2: Pelagic bioluminescence, take 2

Again consider the pelagic bioluminescence data

```{r}
lumin <- read_csv("data/pelagic_bioluminescence.csv") |> clean_names()
```

```{r}
glimpse(lumin)
```

1. Fit an additive model in which a single smooth is used to model all 3 stations, but which the intercept for each station may differ. Print the resulting summary.

2. Plot the observed and predicted bioluminescence versus depth for all 3 stations. Include confidence intervals (hint: you may want to facet your plot)

3. Plot the residuals against depth and use color to distinguish between different stations. Are the model assumptions acceptable?

4. Refit using one smooth for each station. Remember to include all lower-order terms. Repeat parts (1) - (3) for this new fit (i.e., print the summary, make plots of data against predictions and confidence intervals, and investigate model assumptions).

