---
title: "Additive model: basics"
author: "Wing-Ho Ko"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(janitor)
library(mgcv)
library(gratia)
library(broom)
```

## Loading the dataset

```{r}
data(trees, package="datasets")
trees <- trees |> rename(Diam = Girth) |>
  mutate(Hclass = cut(
    Height, breaks=c(-Inf, 70, 80, Inf), labels=c("short", "medium", "tall")
  ), .after=Height) |>
  clean_names()
```

```{r}
glimpse(trees)
```


## Visualization

Visualize the dependence of volume on diameter 

```{r}
ggplot(trees) + geom_point(aes(x=diam, y=volume))
```

Perform the additive model estimation

```{r}
trees_gam <- gam(volume ~ s(diam), gaussian, trees)
```

Print out the summary

```{r}
summary(trees_gam)
```

Using `glance()` and `tidy()` from `broom`

```{r}
glance(trees_gam)
```

```{r}
tidy(trees_gam, parametric=TRUE, conf.int=TRUE)
```

```{r}
tidy(trees_gam)
```

Obtain confidence interval for the smooth

```{r}
trees_diam_confint <- confint(
  trees_gam, 
  "s(diam)", 
  type="confidence", 
  unconditional = TRUE
) |> clean_names()

trees_diam_confint
```

Visualize the result

```{r}
ggplot(
  trees_diam_confint, 
  aes(x=diam, y=estimate, ymax=upper_ci, ymin=lower_ci)
) +
  geom_ribbon(alpha=0.2) +
  geom_line() +
  theme_bw()
```

Add partial residuals:

```{r}
trees_diam_resid <- trees |> 
  add_partial_residuals(trees_gam) |>
  clean_names()

trees_diam_resid
```

Visualize the result again

```{r}
ggplot(
  trees_diam_confint, 
  aes(x=diam, y=estimate)
) +
  geom_ribbon(aes(ymax=upper_ci, ymin=lower_ci), alpha=0.2) +  
  geom_line() +
  geom_point(aes(y=s_diam), data=trees_diam_resid) + 
  theme_bw()
```


Make basic predictions

```{r}
Diam_new <- seq(8, 21, 0.25)
trees_new <- data.frame(Diam=Diam_new)
Volume_new <- predict(trees_gam, trees_new)
trees_new <- cbind(trees_new, data.frame(Volume=Volume_new))

trees_new
```

Make predictions with confidence intervals using `fitted_values()` from `gratia`

```{r}
trees_fitted <- fitted_values(trees_gam, data.frame(Diam=Diam_new), unconditional=TRUE)
trees_fitted
```

Visualize the result

```{r}
ggplot(trees, aes(x=Diam, y=Volume)) + geom_point() + 
  geom_ribbon(
    aes(y=NULL, ymax=.upper_ci, ymin=.lower_ci), 
    data=trees_fitted,
    alpha=0.2
  ) + 
  geom_line(aes(y=.fitted), data=trees_fitted) +
  theme_bw()
```

