---
title: "Additive model: advanced"
author: "Wing-Ho Ko"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(janitor)
library(mgcv)
library(gratia)
library(broom)
```

## Loading the dataset

```{r}
data(trees, package="datasets")
trees <- trees |> rename(Diam = Girth) |>
  mutate(Hclass = cut(
    Height, breaks=c(-Inf, 70, 80, Inf), labels=c("short", "medium", "tall")
  ), .after=Height) |>
  clean_names()
```

```{r}
glimpse(trees)
```


## Extension 1: including parametric terms

```{r}
trees_gam1 <- gam(volume ~ s(diam) + hclass, gaussian, trees)
summary(trees_gam1)
```

As before, we need to treat the `Hclass` factor as a whole, via `anova()`:

```{r}
anova(trees_gam1)
```

Confidence interval for parametric term can be assessed via `tidy()` using `parametric=TRUE` and `conf.int=TRUE`

```{r}
trees_para_confint <- tidy(trees_gam1, parametric=TRUE, conf.int=TRUE)
trees_para_confint
```


## Extension 2: including additional smooths

```{r}
trees_gam2 <- gam(volume ~ s(diam) + s(height), gaussian, trees)
summary(trees_gam2)
```

Extract the predicted smooths and their confidence intervals

```{r}
trees_diam_confint2 <- confint(trees_gam2, "s(diam)", type="confidence") |> 
  clean_names()

trees_height_confint2 <- confint(trees_gam2, "s(height)", type="confidence") |> 
  clean_names()

trees_height_confint2
```

Construct the partial residuals for each variable (in one go)

```{r}
trees_smooth_resid2 <- trees |> add_partial_residuals(trees_gam2) |>
  clean_names()

trees_smooth_resid2
```

Plot smooth with partial residual for diameter

```{r}
ggplot(trees_diam_confint2, aes(x=diam, y=estimate)) +
  geom_point(aes(y=s_diam), data=trees_smooth_resid2) + 
  geom_line() + 
  geom_ribbon(aes(ymax=upper_ci, ymin=lower_ci), alpha=0.2) + 
  theme_bw()
```

Plot smooth with partial residual for height

```{r}
ggplot(trees_height_confint2, aes(x=height, y=estimate)) +
  geom_point(aes(y=s_height), data=trees_smooth_resid2) + 
  geom_line() + 
  geom_ribbon(aes(ymax=upper_ci, ymin=lower_ci), alpha=0.2) + 
  theme_bw()
```

## Extension 3: interactions between parametric and smooth terms

```{r}
ggplot(trees) + geom_point(aes(x=diam, y=volume, color=hclass))
```


```{r}
trees_gam3 <- gam(volume ~ s(diam, by=hclass) + hclass, gaussian, trees)
summary(trees_gam3)
```

Notice that each factor level has a separate smooth. Thus we do not need to include the main effect term `s(diam)` in this case

Construct the prediction with confidence intervals

```{r}
trees_new3 <- rbind(
  data.frame(
    diam=seq(8, 14, 0.25),
    hclass="short"
  ),
  data.frame(
    diam=seq(10, 18, 0.25),
    hclass="medium"
  ),
  data.frame(
    diam=seq(10, 20, 0.25),
    hclass="tall"
  )
) |> mutate(hclass=as.factor(hclass))
```

```{r}
trees_pred3 <- fitted_values(trees_gam3, trees_new3) |> clean_names()
trees_pred3
```

```{r}
ggplot(trees, aes(x=diam, color=hclass, fill=hclass)) + 
  geom_point(aes(y=volume)) + 
  geom_ribbon(aes(ymin=lower_ci, ymax=upper_ci), alpha=0.2, trees_pred3) +
  geom_line(aes(y=fitted), trees_pred3) +
  theme_bw()
```

## Extension 4: multivariable smooths

Since a tree is anisotropic in its dimensions, it makes more sense to use tensor product smooths. And to separate the main and the interaction effects, the following model is fitted:

```{r}
trees_gam4 <- gam(volume ~ s(diam) + s(height) + ti(diam, height), gaussian, trees)
summary(trees_gam4)
```

