---
title: 'Additive model: selection and diagnostics'
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(janitor)
library(mgcv)
library(gratia)
library(broom)
```

## Loading the dataset

```{r}
data(trees, package="datasets")
trees <- trees |> rename(Diam = Girth) |>
  mutate(Hclass = cut(
    Height, breaks=c(-Inf, 70, 80, Inf), labels=c("short", "medium", "tall")
  ), .after=Height) |>
  clean_names()
```

```{r}
glimpse(trees)
```

## Model selection

### Selection of un-penalized terms

Consider the question of whether `hclass` should be included in the model:

```{r}
trees_gam1 <- gam(volume ~ s(diam), gaussian, trees)
trees_gam1$gcv.ubre
```

```{r}
trees_gam2 <- gam(volume ~ s(diam) + hclass, gaussian, trees)
trees_gam2$gcv.ubre
```

So it make sense to include `hclass` in the model.

### Selection of penalized terms

Consider the two-smooths additive model:

```{r}
trees_gam3 <- gam(volume ~ s(diam) + s(height), gaussian, trees)
summary(trees_gam3)
```

We see that the height smooth has an edf close to 1, suggesting that a linear fit may be adaquate. Double check with a plot:

```{r}
trees_height_confint3 <- confint(trees_gam3, "s(height)", type="confidence") |> 
  clean_names()

ggplot(trees_height_confint3, aes(x=height, y=estimate)) +
  geom_line() + 
  geom_ribbon(aes(ymax=upper_ci, ymin=lower_ci), alpha=0.2) + 
  theme_bw()
```

So we may want to fit a model where height is parametric instead:

```{r}
trees_gam4 <- gam(volume ~ s(diam) + height, gaussian, trees)
summary(trees_gam4)
```

## Model diagnostics

Start with the simplest model:

```{r}
trees_gam <- gam(volume ~ s(diam), gaussian, trees)
summary(trees_gam)
```

Use a larger number of `k` and see if there are substantial changes

```{r}
trees_gam <- gam(volume ~ s(diam, k=20), gaussian, trees)
summary(trees_gam)
```

So the default value of `k` is sufficient

Now look at residuals

```{r}
ggplot() + geom_point(aes(x=trees$diam, y=resid(trees_gam)))
```

```{r}
ggplot() + geom_point(aes(x=fitted(trees_gam), y=resid(trees_gam)))
```

There might be heteroscedasticity at the low tree diameter / volume, but with the relatively few data points the result is somewhat inconclusive.

Now let's look at Q-Q plot

```{r}
ggplot(mapping=aes(sample=resid(trees_gam))) + 
  geom_qq() + geom_qq_line()
```

There is no major issue with the Q-Q plot.
